---
title: "Recreating an Edward Tufte chart"
output:
  html_document:
    fig_width: 11
    self_contained: no
date: "Monday, January 05, 2015"
---

The code below recreates the recent charts by [bradleyboehmke](https://twitter.com/bradleyboehmke) ([code](http://t.co/FTLquE5K5p)) and [abresler](https://twitter.com/abresler) ([code](https://gist.github.com/abresler/46c36c1a88c849b94b07)). By using some of the more esoteric features in ggplot2, I was able to reproduce the charts with substantially fewer lines of code.

```{r options, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE, cache = FALSE, 
                      eval = TRUE, cache = FALSE)
theme_set(theme_grey())
```

```{r download_data, eval = FALSE}
lapply(c("ggplot2", "dplyr"), library, character.only = TRUE)
url <- "http://academic.udayton.edu/kissock/http/Weather/gsod95-current/OHDAYTON.txt"
download.file(url, "OHDAYTON.txt")
```

```{r load_data, eval = TRUE}
# Load and clean the data
temp <- read.table("OHDAYTON.txt", na.strings = "-99") %>% 
  setNames(c("month", "day", "year", "temp")) %>% 
  group_by(year) %>% 
  mutate(daynum = seq_along(year))
```

### Plain chart

```{r p_raw}
# Set up plain chart with min-max range and 95% CI
p <- ggplot(temp, aes(x = daynum, y = temp)) + 
  stat_summary(geom = "linerange", 
               fun.ymin = min, 
               fun.ymax = max, 
               color = "wheat2") +
  stat_summary(geom = "linerange", 
               fun.ymin = function(x) mean(x) - 2 * sd(x)/sqrt(length(x)),
               fun.ymax = function(x) mean(x) + 2 * sd(x)/sqrt(length(x)), 
               color = "wheat4") + 
  geom_line(data = filter(temp, year == 2014))

p
```

### Adding points for extreme 2014 temperatures, gridlines, axis labels, etc

```{r df_maxmin}
# Data frame containing all days in 2014 with extreme weather
df_maxmin <- temp %>%
  group_by(daynum) %>%
  mutate(max_temp = max(temp), 
         min_temp = min(temp)) %>%
  filter(year == 2014, 
         temp %in% c(max_temp, min_temp)) %>%
  mutate(max_min = temp == max_temp) # Dummy variable to be mapped to color

head(df_maxmin)
```

```{r df_xaxis}
# Data frame with x-axis breaks and labels
df_xaxis <- temp %>% 
  filter(year == 2014, 
         month != lead(month)) %>%                   # Last day of month
  mutate(days_in_month = daynum - lag(daynum),       # Days in month
         midpoint = lag(daynum) + days_in_month / 2) # Month midpoints
df_xaxis$midpoint[1] <- 31 / 2

head(df_xaxis)
```

```{r p_gridlines}
p <- p  +
  geom_vline(xintercept = 0, color = "wheat4", size = 1) +
  geom_hline(yintercept = seq(-20, 100, 10), color = "white") +
  geom_vline(xintercept = df_xaxis$daynum, 
             color = "wheat4", linetype = "dotted", size = 0.5) +
  geom_point(data = df_maxmin, aes(color = max_min), show_guide = FALSE)

p
```

### Cleaning up scales

```{r p_scales}
p <- p +
  scale_x_continuous(expand = c(0,0), labels = month.name,
                     breaks = c(df_xaxis$midpoint, df_xaxis$daynum[11] + (31/2))) +
  scale_y_continuous(expand = c(0,0), breaks = seq(-20, 100, 10),
                     labels = function(x) parse(text = paste0(x, "*degree"))) +
  scale_color_manual(values = c("blue3", "firebrick3"))

p
```

### Removing chartjunk

```{r p_design}
p <- p + theme(axis.ticks = element_blank(), 
               panel.grid = element_blank(),
               panel.background = element_blank(),
               panel.border = element_blank(),
               axis.text = element_text(color = "gray30"),
               plot.title = element_text(face = "bold", hjust = 0.012, 
                                         vjust = 0.8, color = "#3C3C3C", 
                                         size = 20)) +
                 labs(x = NULL, y = NULL, title = "Dayton's Weather in 2014")

p
```

### Adding description and text annotations

```{r p_titles}
desc <- "Data represents average daily temperatures. Accessible data dates back 
         to January 1, 1975. Data for 2014 is only available through December 16. 
         Average temperature for the year was 51.9° making 2014 the 9th coldest 
         year since 1995." %>% 
  strwrap(width = 0.9 * getOption("width")) %>% 
  paste0(collapse = "\n")

# Data frame with annotations
df_annotate <- data_frame(
  daynum = c(40, 307), temp = c(-10, 82), max_min = c(FALSE, TRUE), 
  label = c("We had 35 days that were the\ncoldest since 1995", 
            "We had 19 days that were\nthe hottest since 1995"))

p <- p + 
  annotate("text", x = 5, y = 96, size = 4, fontface = "bold", 
             hjust = 0, vjust = 0, label = "Temperature") +
  annotate("text", x = 5, y = 94, size = 3, color = "gray30", 
           hjust = 0, vjust = 1, label = desc) +
  geom_segment(aes(x = c(30, 302), xend = c(40, 307), 
                   y = c(-5, 74), yend = c(-10, 82),
                   color = c(FALSE, TRUE)), show_guide = FALSE) + 
  geom_text(data = df_annotate, aes(color = max_min, label = label), 
            size = 3, hjust = 0, show_guide = FALSE)

p
```

### Final figure

```{r p_annotated, eval = TRUE}
# Data frame with legend label coordinates
df_leg_text <- data_frame(daynum = c(186, 145, 184, 184), 
                          temp = c(14.75, 14.75, 25, 5), 
                          label = c("NORMAL RANGE", "2014 TEMPERATURE", 
                                    "RECORD HIGH", "RECORD LOW"))

# Data frame with legend shape coordinates
df_leg_seg <- data_frame(daynum = c(181, 181, 183, 183, 185), 
                         xend   = c(181, 181, 185, 185, 185),
                         temp   = c(5, 12, 18, 12, 12),
                         yend   = c(25, 18, 18, 12, 18), 
                         size   = c(3, 3, 0.5, 0.5, 0.5), 
                         color  = c("wheat2", rep("wheat4", 4)))

p + 
  geom_segment(data = df_leg_seg, aes(xend = xend, yend = yend), 
               size = df_leg_seg$size, color = df_leg_seg$color) +
  geom_line(data = data_frame(daynum = seq(175, 182), temp = rnorm(8, 15, 2))) +
  geom_text(data = df_leg_text, aes(label = label), hjust = 0, size = 2)

p
```
