---
title: "Fun with floats"
output:
  html_document: default
date: "May 25, 2016"
---
  
```{r}
suppressPackageStartupMessages(library(purrr))

with_expr_and_result <- function(f) {
  stopifnot(is_character(f))
  force(f)
  function(...) {
    args <- list(...)
    FUN <- eval(as.name(f))
    x <- FUN(...)
    message(sprintf("(%s %s %s) == %s", args[[1]], f, args[[2]], x))
    x
  }
}

like_ninja <- function(f) {
  function(...) {
    x <- f(...)
    invisible(x)
  }
}

float_equal <- function(e1, e2, tolerance = .Machine$double.eps ^ 0.5) {
  e1 - e2 < tolerance
}

funs <- c("==", "-") %>% map(with_expr_and_result)

f <- function(x) x ^ 2
g <- function(x) sqrt(x)
fog <- function(x) f(g(x))  # fog and gof should return same result
gof <- function(x) g(f(x))

(y <- c(fog, gof) %>% map(~.x(1L:10L)))            # All equal, as expected
y %>% like_ninja(pmap_lgl)(funs[[1]])              # Wait, what? [^1]
y %>% like_ninja(pmap_dbl)(funs[[2]])              # Oh, right. Floats...
y %>% pmap_lgl(float_equal)                        # There we go

lift(all.equal)(y)    # Canonical way to test for equality of floating points

# 1: Or for the sadists: funs %>% like_ninja(map)(~pmap(y, .x))
```

We can use the `Rmpfr` package to investigate further how (un)precise the 
results are:
  
```{r}
suppressPackageStartupMessages(library("Rmpfr"))
y %>% map(mpfr, precBits = .Machine$double.digits)
```

Unsurprisingly, the problem is with the `fog` function, which first takes the
square root and then exponentiates the result.

This result relates back to a comment that Hadley Wickham made on Twitter some
time ago:
  
<blockquote class="twitter-tweet" data-lang="en">
<p lang="en" dir="ltr">
<a href="https://twitter.com/hadleywickham">@hadleywickham</a> Neat! 
<a href="https://t.co/PmJXnYqM45">pic.twitter.com/PmJXnYqM45</a>
</p>&mdash; EP (@expersso) 
<a href="https://twitter.com/expersso/status/676771253647253504">December 15, 2015</a>
</blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
  
```{r}
wat <- function(x) (1 / x) * x
wat(1:49) %>% mpfr(.Machine$double.digits)
```

